---
output: github_document
---

<!-- index.md is generated from index.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

## Introducing dominoSignal <img src="logo.svg" alt="dominoSignal hex logo" align="right" height="138" style="float:right; height:138px;" />

dominoSignal infers intra- and intercellular signaling from single cell RNA sequencing (scRNAseq) data. Within each cell cluster, it identifies transcription factors (TFs) with enriched activity and links them to receptors whose expression correlates with that activity. It then connects those active receptors to ligands expressed by other clusters, producing a cluster-to-cluster signaling network associated with downstream transcriptional response rather than ligand and receptor expression alone.

dominoSignal builds on the original [domino](https://github.com/Elisseeff-Lab/domino) R package ([Cherry et al., 2021, *Nature Biomedical Engineering*](https://doi.org/10.1038/s41551-021-00770-5)) and adds the Differential Cell Signaling Test (DCST) for statistically comparing signaling across subjects and conditions ([Mitchell et al., 2026, *Bioinformatics*](https://doi.org/10.1093/bioinformatics/btag089)).

## Installation

dominoSignal is available from [Bioconductor](https://bioconductor.org/packages/release/bioc/html/dominoSignal.html):

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("dominoSignal")
```

The development version, with the latest features and fixes, is available from Bioconductor devel:

```{r, eval = FALSE}
BiocManager::install(version = "devel")
BiocManager::install("dominoSignal")
```

Changes between versions, including any that affect results, are listed in the [changelog](https://fertiglab.github.io/dominoSignal/news/index.html).

## Workflow

### 1. Prepare inputs

dominoSignal needs four inputs:

- **Expression data**: a counts matrix, a z-scored (scaled) expression matrix, and a named factor of cluster labels, with cell names matching across all three. These can be extracted from a `SingleCellExperiment` or similar object.
- **TF activity scores**: a matrix of TF activity per cell, plus an optional list of each TF's target genes. We recommend [pySCENIC](https://pyscenic.readthedocs.io/en/latest/); `create_regulon_list_scenic()` converts its regulon output. See [SCENIC for TF Activation Scoring](https://fertiglab.github.io/dominoSignal/articles/tf_scenic_vignette.html).
- **Ligand-receptor database**: a data frame describing receptor-ligand pairs. We recommend [CellPhoneDB](https://www.cellphonedb.org/); `create_rl_map_cellphonedb()` builds this table from its database files. See [Using the CellPhoneDB Database](https://fertiglab.github.io/dominoSignal/articles/cellphonedb_vignette.html).

Other TF scoring methods and ligand-receptor databases can be used, provided they are provided to `create_domino()` in the correct format.

### 2. Build a signaling network

`create_domino()` stores the inputs, tests TF enrichment by cluster, and correlates receptor expression with TF activity. `build_domino()` then applies thresholds to assemble the signaling network.

```{r, eval = FALSE}
library(dominoSignal)

dom <- create_domino(
    rl_map = rl_map, features = tf_scores, counts = counts,
    z_scores = z_scores, clusters = clusters, tf_targets = regulon_list
)
dom <- build_domino(
    dom,
    max_tf_pval = 0.001, max_tf_per_clust = 25, max_rec_per_tf = 25,
    rec_tf_cor_threshold = 0.25, min_rec_percentage = 0.1
)
```

See [Get Started with dominoSignal](https://fertiglab.github.io/dominoSignal/articles/dominoSignal.html) for a complete example, including guidance on choosing these thresholds.

### 3. Explore and visualize results

Results are stored in a domino object. Accessor functions (`dom_signaling()`, `dom_linkages()`, `dom_info()`, and others) retrieve its contents, and `dom_to_df()` returns the signaling results as a data frame. Plotting functions include signaling and gene networks, heatmaps of signaling, TF activity, and receptor-TF correlation, and chord diagrams of ligand expression. See [Interacting with domino Objects](https://fertiglab.github.io/dominoSignal/articles/domino_object_vignette.html) and [Plotting Functions and Options](https://fertiglab.github.io/dominoSignal/articles/plotting_vignette.html).

### 4. Compare signaling across subjects

With a domino object for each subject or sample, the DCST workflow identifies linkages (for example, active receptors or ligand-receptor pairs in a cluster) that differ between groups.

```{r, eval = FALSE}
link_summary <- summarize_linkages(dom_list, subject_meta, subject_names = subject_meta$subject)
diff_links <- test_differential_linkages(
    link_summary, cluster = "CD8_T_cell", group.by = "condition", linkage = "rec"
)
plot_differential_linkages(diff_links, "p.value")
```

See [Differential Signaling Workflow](https://fertiglab.github.io/dominoSignal/articles/differential_signaling.html) for a full walkthrough.

## Citation

If you use dominoSignal, please cite:

> Cherry C, Maestas DR, Han J, Andorko JI, Cahan P, Fertig EJ, Garmire LX, Elisseeff JH. Computational reconstruction of the signalling networks surrounding implanted biomaterials from single-cell transcriptomics. *Nat Biomed Eng*. 2021;5(10):1228-1238. [doi:10.1038/s41551-021-00770-5](https://doi.org/10.1038/s41551-021-00770-5)

If you use the differential signaling workflow, please also cite:

> Mitchell JT, Stapleton O, Krishnan K, Nagaraj S, Lvovs D, Cherry C, Poissonnier A, Horton W, Adey A, Rao V, Huff A, Zimmerman JW, Kagohara LT, Zaidi N, Coussens LM, Jaffee EM, Elisseeff JH, Fertig EJ. Differential cell signaling testing for cell-cell communication inference from single-cell data by dominoSignal. *Bioinformatics*. 2026;42(3):btag089. [doi:10.1093/bioinformatics/btag089](https://doi.org/10.1093/bioinformatics/btag089)

## Getting help

If you find a bug or have a question, please [open an issue](https://github.com/FertigLab/dominoSignal/issues).
