---
title: "Differential Signaling Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differential Signaling Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.cap = "",
  tidy = TRUE
)
options(timeout = 600)
```

In this tutorial, we will walk through the workflow for performing differential signaling analysis using the dominoSignal package. While the other tutorials focus on signaling networks inferred from a single data set, many studies involve multiple subjects or conditions, requiring the comparison of signaling across these groups. The dominoSignal package implements the Differential Cell Signaling Test (DCST), a framework for identifying receptor, ligand, and transcription factor (TF) linkages that are significantly differentially present across groups of subjects. This method is described in detail in [Differential cell signaling testing for cell-cell communication inference from single-cell data by dominoSignal](https://doi.org/10.1093/bioinformatics/btag089) (2026) published in Bioinformatics. The publication also includes simulation benchmarking and application of the method to pancreatic cancer and immunotherapy data sets. Refer to the publication for full details on the method and biological validation of the workflow.

## Workflow overview

The differential signaling workflow consists of the following steps:

1. Build a domino object per subject to be compared (see the [Getting Started](https://fertiglab.github.io/dominoSignal/articles/dominoSignal.html) tutorial for details on building a domino object).
2. Compile linkages across the subjects into a [linkage_summary object](https://fertiglab.github.io/dominoSignal/reference/linkage_summary-class.html) using the `summarize_linkages()` function.
3. [Optional] Subset the `linkage_summary` object to include only the subjects of interest.
4. Create a table of linkage observations within groups of subjects using the `count_linkage()` function.
5. Test differences in linkage frequency across groups using the `test_differential_linkages()` function, which implements a Fisher's exact test.
6. Visualize the results using the `plot_differential_linkages()` function.

The DCST workflow can be performed with 2 or more groups of subjects. In this tutorial, we will demonstrate the workflow for both a two-group comparison (condition A vs condition B) and a comparison across more than two groups.

## Demonstration data

In this tutorial, we use the domino object built in the [Getting Started](https://fertiglab.github.io/dominoSignal/articles/dominoSignal.html) tutorial. If you have not yet built a domino object, you can do so by following the instructions on that page or use the code below to load the pre-built domino object.

<details><summary>Instructions to load data</summary>

```{r setup}
set.seed(42)
library(dominoSignal)
library(dplyr)
library(withr)
library(kableExtra)

# BiocFileCache helps with file management across sessions
bfc <- BiocFileCache::BiocFileCache(ask = FALSE)
data_url <- "https://zenodo.org/records/10951634/files/pbmc_domino_built.rds"
tmp_path <- BiocFileCache::bfcrpath(bfc, data_url)
dom <- readRDS(tmp_path)
```

</details>

Because the differential signaling workflow compares linkages across multiple domino objects (one per subject), we will create synthetic domino objects for four additional subjects from the same PBMC dataset. For each synthetic subject, we randomly retain a subset of the cells without replacement while keeping their true cluster labels and then perturb the cluster labels of a subset of those cells by reshuffling within that subset. This approach preserves some amount of per-cluster signaling while making sure each domino object is distinct from the others, giving the example workflow something to compare.^[Other methods, such as shuffling all cluster labels, results in clusters that are all statistically similar which results in few linkages reaching significance. While this approach reduces the differences between domino objects, it preserves enough signaling to infer a reasonable number of (insignificant) linkages.]

Note that reshuffling is only for demonstration purposes and does not represent any biological differences. In practice, you would build one domino object from each biological subject or sample to be used in the comparison. To ensure results are comparable across subjects, use the same parameters in `create_domino()` and `build_domino()` for all subjects. Refer to the [DCST publication](https://doi.org/10.1093/bioinformatics/btag089) for an example of this workflow applied to real data.

<details><summary>Creating synthetic replicates</summary>

To start, we extract the inputs from the original domino object so we can supply them to the `create_domino()` function for the synthetic subjects.
```{r extract inputs}
rl_map <- dom_database(dom, name_only = FALSE)
features <- dom_tf_activation(dom)
# Access counts slot directly to maintain sparse matrix format
counts <- slot(dom, "counts")
zscores <- dom_zscores(dom)
tf_targets <- dom_linkages(dom, link_type = "tf-target")
clusters <- dom_clusters(dom, labels = TRUE)
```

We can use a helper function to resample the cells and shuffle the cluster labels for a subset of them, and then rebuild the domino object with `create_domino()` and `build_domino()` using the same parameters across all subjects so that they can be compared.^[These parameters are fairly permissive to ensure enough linkages are inferred for demonstration purposes.]

```{r create replicates, result = "hide"}
build_reps <- function(clusters, seed, keep_frac, shuffle_frac) {
  # Local seed for reproducibility without affecting global seed
  withr::with_seed(seed, {
    n_cells <- length(clusters)
    keep_cells <- sort(sample(seq_len(n_cells), size = round(n_cells * keep_frac), replace = FALSE))

    resampled_clusters <- clusters[keep_cells]
    shuffled_idx <- sample(seq_along(resampled_clusters), size = round(shuffle_frac * length(resampled_clusters)), replace = FALSE)
    
    resampled_clusters[shuffled_idx] <- sample(resampled_clusters[shuffled_idx])

    dom_shuffle <- create_domino(
      rl_map = rl_map,
      features = features,
      counts = counts,
      z_scores = zscores,
      clusters = resampled_clusters,
      tf_targets = tf_targets,
      use_clusters = TRUE,
      use_complexes = TRUE,
      remove_rec_dropout = FALSE,
      verbose = FALSE
    )

    dom_shuffle <- build_domino(
      dom = dom_shuffle,
      min_tf_pval = 0.05,
      max_tf_per_clust = 25,
      max_rec_per_tf = 25,
      rec_tf_cor_threshold = 0.1,
      min_rec_percentage = 0.05)

  })

  return(dom_shuffle)
}

params <- data.frame(set = c(161803, 314159, 6631034, 6021023, 831446, 98067),
    keep = c(0.95, 0.95, 0.95, 0.95, 0.9, 0.9),
    shuffle = c(0.15, 0.1, 0.15, 0.1, 0.15, 0.1))

# This does take a few minutes to run
dom_list <- purrr::pmap(params, function(set, keep, shuffle) {
  build_reps(clusters, seed = set, keep_frac = keep, shuffle_frac = shuffle)
})

names(dom_list) <- paste0("P", seq_along(dom_list))
```

We now have six domino objects that we can treat as different subjects (P1-6).

</details>

## The `linkage_summary` class

The [linkage_summary class](https://fertiglab.github.io/dominoSignal/reference/linkage_summary-class.html) is the structure used for the differential signaling workflow. It contains linkages present in each cluster across the domino objects it contains, along with subject metadata. The object contains three slots:

- `subject_names`: a factor with a unique name for each subject (domino object) represented in the summary.
- `subject_meta`: a data frame with one row per subject including provided metadata (such as condition, treatment, or other covariates) that can be used to group or subset subjects for comparison.
- `subject_linkages`: a nested list organized as subject to cluster to linkage type (`tfs` for active transcription factors, `rec` for active receptors, `incoming_lig` for ligands that can activate the cluster, `tfs_rec` for transcription factor to receptor pairs, and `rec_lig` for receptor to ligand pairs).

To create a `linkage_summary` object, we use the `summarize_linkages()` function, which takes a named list of domino objects and a data frame of subject metadata. The first column of the data frame must contain subject names that match the names of the domino objects in the list.

```{r summarize linkages, message = TRUE}

subject_metas <- data.frame(
  patient = c("P1", "P2", "P3", "P4", "P5", "P6"),
  time = c("pre", "post", "pre", "post", "pre", "post"),
  condition = c("control", "treatment", "treatment", "control", "control", "control"),
  batch = c("A", "A", "B", "B", "C", "C"))

link_summary <- summarize_linkages(dom_list, subject_metas,
    subject_names = subject_metas$patient)

print(link_summary)
```

The `print` and `show` methods for the `linkage_summary` class report the number of subjects, the number of metadata annotations, and the number of clusters present in the summary.

Since we built this linkage summary from six domino objects with six distinct groups and two distinct conditions, we can subset the summary to include only the subjects of interest. 

To create a comparison for only the treatment condition, we can subset the `linkage_summary` object by the relevant metadata column.

```{r subset treatment, message = TRUE}
link_treat <- subset(link_summary, subset = condition == "treatment")
print(link_treat)
```

Subjects can also be selected by name or with compound expressions.

```{r subset examples, message = TRUE}
subset(link_summary, subset = subject_names %in% c("P1", "P3"))
subset(link_summary, subset = condition == "treatment" | time == "pre")
```

As an alternative to creating separate subset objects, the other functions in the differential signaling workflow accept a `subject_names` argument that can be used to restrict the subjects considered in the analysis without needing to create a separate `linkage_summary` object.

## Counting linkages across groups

The `count_linkage()` function tabulates the number of subjects with a given linkage in a specified cluster, which can also be broken down by a metadata column. The `linkage` argument selects which of the linkage types to count.

### Total counts

If no grouping variable is specified and no subject names are provided, the function will return a data frame with the total number of subjects with each linkage in the selected cluster.

```{r all linkages}
count_all <- count_linkage(link_summary, cluster = "CD8_T_cell",
    linkage = "rec_lig") %>% slice_sample(n = 5)
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(count_all,
    caption = "Total counts of receptor-ligand linkages in CD8 T Cells"),
    bootstrap_options = c("striped", "hover"))
```
</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(count_all, caption = "Total counts of receptor-ligand linkages in CD8 T Cells"), bootstrap_options = c("striped", "hover"))
```

If specific subjects are of interest, the counts can be restricted to those subjects by providing a vector of the subject names.

```{r subset subjects}
count_subjects <- count_linkage(link_summary, cluster = "CD8_T_cell",
    linkage = "tfs", subject_names = c("P1", "P2"))
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(count_subjects,
    caption = "Counts of TF linkages in patients P1 & P2 in CD8 T Cells"),
    bootstrap_options = c("striped", "hover"))
```
</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(count_subjects, caption = "Counts of TF linkages in patients P1 & P2 in CD8 T Cells"), bootstrap_options = c("striped", "hover"))
```

### Grouped counts

The `count_linkage()` function can also return counts that are grouped by a metadata column. For example, we can count the linkages for subjects in the treatment condition grouped by the "time" column to see how many subjects in each time point have each linkage.

```{r time treatment counts}
counts_time_trt <- count_linkage(link_treat, cluster = "CD14_monocyte",
    linkage = "tfs_rec", group.by = "time")
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(slice_sample(counts_time_trt, n = 5),
    caption = "Counts of TF-Receptor linkages in CD14 Monocytes grouped by time for treatment subjects"),
    bootstrap_options = c("striped", "hover"))
```
</details>

```{r include = FALSE}
kableExtra::kable_styling(knitr::kable(slice_sample(counts_time_trt, n = 5),
    caption = "Counts of TF-Receptor linkages in CD14 Monocytes grouped by time for treatment subjects"),
    bootstrap_options = c("striped", "hover"))
```

This is equivalent to grouping by the "time" column and providing the subject names of only the subjects in the treatment condition.

```{r time subject counts}
counts_time_subj <- count_linkage(link_summary, cluster = "CD14_monocyte",
    linkage = "tfs_rec", group.by = "time", subject_names = c("P2", "P3"))
identical(counts_time_trt, counts_time_subj)
```

Of course, more than one subject can be in each group, and the counts will reflect the number of subjects in each group with each linkage.

```{r condition counts}
counts_condition <- count_linkage(link_summary, cluster = "naive_CD4_T_cell",
    linkage = "incoming_lig", group.by = "condition") %>% slice_sample(n = 5)
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(counts_condition,
    caption = "Counts of incoming ligand linkages in naive CD4 T Cells grouped by condition"),
    bootstrap_options = c("striped", "hover"))
```
</details>


```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(counts_condition, caption = "Counts of incoming ligand linkages in naive CD4 T Cells grouped by condition"), bootstrap_options = c("striped", "hover"))
```

And more than two groups can be compared at once.

```{r batch counts}
counts_batch <- count_linkage(link_summary, cluster = "CD14_monocyte",
    linkage = "rec", group.by = "batch") %>% slice_sample(n = 5)
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(counts_batch,
    caption = "Counts of receptor linkages in CD14 Monocytes grouped by batch"),
    bootstrap_options = c("striped", "hover"))
```
</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(counts_batch, caption = "Counts of receptor linkages in CD14 Monocytes grouped by batch"), bootstrap_options = c("striped", "hover"))
```

## Testing differential linkages

The `count_linkage()` function tabulates the number of subjects with a given linkage in a specified cluster, but it does not test whether the frequency of that linkage is significantly different across groups. The `test_differential_linkages()` function uses a contingency table of presence/absence of each linkage across groups to perform a Fisher's exact test. The p-values across all tested linkages are adjusted for multiple comparisons using the false discovery rate method.

The `test_differential_linkages()` function returns a data frame that includes the cluster, linkage, grouping variable, test name (always fisher's exact test), the linkage being compared, the test statistics (odds ratio, if applicable, and p-value, both raw and adjusted), total number of subjects where linkage is active, number of subjects in each group where linkage is active, number of total subjects compared, and the total number of subjects in each group.

### Two-group comparisons

When exactly two groups are compared, the `fisher.test()` function returns an odds ratio in addition to the p-value, reflecting the relative odds of the linkage being present in one group compared to the other.

```{r test time treat}
dcst_time_trt <- test_differential_linkages(link_treat,
    cluster = "CD14_monocyte", linkage = "tfs_rec", group.by = "time")
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(slice_sample(dcst_time_trt, n = 5),
    caption = "DCST of TF-Receptor linkages in CD14 Monocytes grouped by time for treatment subjects"),
    bootstrap_options = c("striped", "hover"))
```
</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(slice_sample(dcst_time_trt, n = 5), caption = "DCST of TF-Receptor linkages in CD14 Monocytes grouped by time for treatment subjects"), bootstrap_options = c("striped", "hover"))
```

As with the `count_linkage()` function, the same results can be obtained by providing the subject names of only the subjects in the treatment condition.

```{r test time subject}
dcst_time_subj <- test_differential_linkages(link_summary, cluster = "CD14_monocyte",
    linkage = "tfs_rec", group.by = "time", subject_names = c("P2", "P3"))
identical(dcst_time_trt, dcst_time_subj)
```

The `test_differential_linkages()` function does require that any grouping variable has at least two groups with at least one subject in each group.

```{r no subjects, error = TRUE}
test_differential_linkages(link_treat, cluster = "CD14_monocyte",
    linkage = "tfs_rec", group.by = "condition")
```

It is obviously preferable to have more than one subject in each group (though the function does return results for groups with only one subject). Note that there are no significant differences between the groups in this example, which is expected since the subjects were generated from the same data set.

```{r dcst time}
dcst_time <- test_differential_linkages(link_summary, cluster = "CD8_T_cell",
    linkage = "rec", group.by = "time") %>% dplyr::arrange(desc(odds.ratio))
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(slice_sample(dcst_time, n = 5),
    caption = "DCST of receptor-ligand linkages in CD8 T Cells grouped by time"),
    bootstrap_options = c("striped", "hover"))
```
</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(slice_sample(dcst_time, n = 5), caption = "DCST of receptor-ligand linkages in CD8 T Cells grouped by time"), bootstrap_options = c("striped", "hover"))
```

### Multi-group comparisons

When more than two groups are compared, the `fisher.test()` function does not return an odds ratio (so it is set to `NA` in the DCST output), but the p-value still reflects whether the frequency of the linkage is significantly different across groups.

```{r dcst batch}
dcst_batch <- test_differential_linkages(link_summary, cluster = "naive_CD4_T_cell",
    linkage = "rec", group.by = "batch") %>% arrange(p.value)
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(slice_sample(dcst_batch, n = 5),
    caption = "DCST of receptors in naive CD4 T Cells grouped by batch"),
    bootstrap_options = c("striped", "hover"))
```
</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(slice_sample(dcst_batch, n = 5), caption = "DCST of receptors in naive CD4 T Cells grouped by batch"), bootstrap_options = c("striped", "hover"))
```

There is no limit to the number of groups that can be compared at once (as long as there is at least one subject in each group).

```{r dcst patient}
dcst_patient <- test_differential_linkages(link_summary, cluster = "naive_CD4_T_cell",
    linkage = "rec_lig", group.by = "patient") %>% arrange(p.value)
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(slice_sample(dcst_patient, n = 5),
    caption = "DCST of receptor-ligand linkages in naive CD4 T Cells grouped by patient"),
    bootstrap_options = c("striped", "hover"))
```
</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(slice_sample(dcst_patient, n = 5), caption = "DCST of receptor-ligand linkages in naive CD4 T Cells grouped by patient"), bootstrap_options = c("striped", "hover"))
```

The [DCST paper](https://doi.org/10.1093/bioinformatics/btag089) includes additional information on the recommended number of subjects for robust results (at least 15 per group), as well as an approach to simulate biological replicates using bootstrapping of cells for experiments with small numbers of replicates (scripts for the manuscript can be found at [the corresponding GitHub repository](https://github.com/FertigLab/Differential_Cell_Signaling_Test)).

## Visualizing differential linkages

In order to visualize the results of the DCST, the `plot_differential_linkages()` function can be used to create a heatmap of the test statistic along with row-level bar plots showing the proportions of subjects in which linkages are active.

The function has arguments for the test statistic (the column of the DCST result that contains the statistic to be plotted), the range of values to be plotted, the heatmap gradient, and the color palette to use for the groups being compared.

### Two-group comparisons

An example of a two-group comparison is shown below, where the p-value is used as the test statistic and the color scale is set to ascending so that the lower p-values are darker (resulting in a heatmap where all cells are white since the p-values are 1 for all linkages in this example).

```{r plot time treat}
#| fig.alt: Heatmap of p-values for receptor-ligand linkages in CD8 T cell cluster differing between pre and post timepoint for treatment condition with row barplots showing proportion of subjects in each group with active linkage.
plot_differential_linkages(differential_linkage = dcst_time_trt %>% slice_sample(n = 15),
  test_statistic = "p.value", stat_range = c(0, 1), stat_ranking = "ascending")
```

```{r plot time}
#| fig.alt: Heatmap of p-values for receptors in CD8 T cells differing between pre and post timepoint with row barplots showing proportion of subjects in each group with active receptor.
plot_differential_linkages(differential_linkage = dcst_time %>% slice_sample(n = 15),
    test_statistic = "p.value", gradient = c(minimum = "#c44f4f", maximum = "#ffd771"),
    stat_ranking = "ascending")
```

If comparing between two groups, it is also possible to use the odds ratio as the test statistic, which will show the relative odds of a linkage being present in one group compared to the other.

```{r plot dcst time odds}
#| fig.alt: Heatmap of odds ratios for receptors in CD8 T cells differing between pre and post timepoints with row barplots showing proportion of subjects in each group with active receptor.
plot_differential_linkages(differential_linkage = dcst_time %>% slice_sample(n = 15),
    test_statistic = "odds.ratio", stat_range = c(0, 4), stat_ranking = "descending",
    gradient = c(minimum = "gray90", maximum = "orange"))
```

### Multi-group comparisons

When comparing between more than two groups, the odds ratio is not available, so the p-value (or adjusted p-value) is the only option for the test statistic.

```{r plot dcst batch}
#| fig.alt: Heatmap of p-values for receptors in naive CD4 T cells differing between batches with row barplots showing proportion of subjects in each group with active receptor.
plot_differential_linkages(differential_linkage = dcst_batch %>% slice_sample(n = 15),
  test_statistic = "p.value", stat_range = c(0, 1), stat_ranking = "ascending",
  gradient = c(minimum = "#ff4c43", maximum = "#ffe6c0"))
```

A custom color can be assigned to each group's barplot annotation using the `group_palette` argument, which takes a named vector of colors where names correspond to the group names in the DCST result.

```{r plot dcst group colors}
plot_differential_linkages(differential_linkage = dcst_batch %>% slice_sample(n = 15),
    test_statistic = "p.value", stat_range = c(0, 1), stat_ranking = "ascending",
    gradient = c(minimum = "#7b43ff", maximum = "#c0c5ff"),
    group_palette = c(A = "#64a5e6", B = "#71fcad", C = "#c968c3"))
```

This can be useful for consistency across multiple plots or to match publication color schemes.

## Example run

Here's an example of what a workflow might look like from the list of `domino()` objects to the final visualization of differential linkages. We begin with the list of synthetic domino objects created above, create the metadata, create a `link_summary()` object, examine the counts for the linkages,and then run the DCST for a comparison by batch. In this example, we focus on incoming ligands in the CD14 monocyte cluster.

```{r example workflow, eval = FALSE}
subject_metas <- data.frame(
  patient = c("P1", "P2", "P3", "P4", "P5", "P6"),
  time = c("pre", "post", "pre", "post", "pre", "post"),
  condition = c("control", "treatment", "treatment", "control", "control", "control"),
  batch = c("A", "A", "B", "B", "C", "C"))

link_summary <- summarize_linkages(dom_list, subject_metas,
    subject_names = subject_metas$patient)

counts_batch <- count_linkage(link_summary, cluster = "CD14_monocyte",
    linkage = "incoming_lig", group.by = "batch")

dcst_batch <- test_differential_linkages(link_summary, cluster = "CD14_monocyte",
    linkage = "incoming_lig", group.by = "batch")

plot_differential_linkages(dcst_batch, test_statistic = "p.value", stat_ranking = "ascending", 
    gradient = c("minimum" = "#e02c2c", "maximum" = "#feffcd"),
    group_palette = c(A = "#2a6aaa", B = "#07bd56", C = "#a244a3"))

```

## Continued development

As dominoSignal is actively being developed, new functions and features may be implemented in future versions. In the meantime, see our [Getting Started](https://fertiglab.github.io/dominoSignal/articles/dominoSignal.html) tutorial for the basic workflow for building a domino object and inferring signaling networks, the [domino object structure](https://fertiglab.github.io/dominoSignal/articles/domino_object_vignette.html) vignette to get familiar with the domino object structure, or our [plotting vignette](https://fertiglab.github.io/dominoSignal/articles/plotting_vignette.html) for visualization options for single subject domino objects. If you find any bugs, have questions, or would like to suggest new features, please submit an issue on the [dominoSignal GitHub repository](https://github.com/FertigLab/dominoSignal/issues).


<details><summary>Vignette Build Information</summary>
Date last built and session information:

```{r}
Sys.Date()
sessionInfo()
```

</details>