---
title: "Get Started with dominoSignal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with dominoSignal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

dominoSignal is a tool for analysis of intra- and intercellular signaling in single cell RNA sequencing (scRNAseq) data based on transcription factor (TF) activity. Here, we show a basic example pipeline for generating communication networks as a `domino()` object. This vignette will demonstrate how to use dominoSignal on the 10X Genomics Peripheral Blood Mononuclear Cells (PBMC) data set of 2,700 cells. The data can be downloaded [from 10x genomics](https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz). This vignette uses [a preprocessed version from Zenodo](https://doi.org/10.5281/zenodo.10124865).


## Installation

Installation of dominoSignal from Bioconductor can be achieved using the `{BiocManager}` package.

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE)){
    install.packages("BiocManager")
}

BiocManager::install("dominoSignal")
```


## Options and Setup
```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.cap = "",
    tidy = TRUE
)
options(timeout = 300)
```

<details><summary>Libraries and set up</summary>

```{r libraries}
set.seed(42)
library(dominoSignal)
library(SingleCellExperiment)
library(knitr)
```

</details>

<details><summary>Data used in our vignettes can be downloaded from Zenodo.</summary>

```{r files}
# BiocFileCache helps with managing files across sessions
bfc <- BiocFileCache::BiocFileCache(ask = FALSE)
data_url <- "https://zenodo.org/records/10951634/files"

# download the reduced pbmc files
# preprocessed PBMC 3K scRNAseq data set as a SingleCellExperiment object
pbmc_url <- paste0(data_url, "/pbmc3k_sce.rds")
pbmc <- BiocFileCache::bfcrpath(bfc, pbmc_url)

# download scenic results
scenic_auc_url <- paste0(data_url, "/auc_pbmc_3k.csv")
scenic_auc <- BiocFileCache::bfcrpath(bfc, scenic_auc_url)

scenic_regulon_url <- paste0(data_url, "/regulons_pbmc_3k.csv")
scenic_regulon <- BiocFileCache::bfcrpath(bfc, scenic_regulon_url)

# download CellPhoneDB files
cellphone_url <- "https://github.com/ventolab/cellphonedb-data/archive/refs/tags/v4.0.0.tar.gz"
cellphone_tar <- BiocFileCache::bfcrpath(bfc, cellphone_url)
cellphone_dir <- paste0(tempdir(), "/cellphone")
untar(tarfile = cellphone_tar, exdir = cellphone_dir)
cellphone_data <- paste0(cellphone_dir, "/cellphonedb-data-4.0.0/data")


# directory for created inputs to pySCENIC and dominoSignal
input_dir <- paste0(tempdir(), "/inputs")
if (!dir.exists(input_dir)) {
    dir.create(input_dir)
}
```

</details>

## Data preparation

Analysis of cell-cell communication with dominoSignal often follows initial processing and annotation of scRNAseq data.  We used the [OSCA workflow](https://bioconductor.org/books/3.18/OSCA.workflows/unfiltered-human-pbmcs-10x-genomics.html) to prepare the data set for analysis with dominoSignal. The complete processing script is available in the `data-raw` directory of the dominoSignal package. The processed data can be downloaded from [Zenodo](https://doi.org/10.5281/zenodo.10124865).

```{r}
pbmc <- readRDS(pbmc)
```


## Loading TF and receptor-ligand data

### Loading SCENIC results

The TF activities are a required input for dominoSignal, and the regulons learned by [SCENIC](https://scenic.aertslab.org/) are an important but optional input needed to annotate TF-target interactions and to prune TF-receptor linkages where the receptor is a target of the TF, as correlation cannot distinguish between the receptor driving TF activity or TF activity inducing receptor expression.

```{r Load SCENIC Results}
regulons <- read.csv(scenic_regulon)
auc <- read.table(scenic_auc,
    header = TRUE, row.names = 1,
    stringsAsFactors = FALSE, sep = ","
)
```

The initial regulons data frame read into R from the ctx function has two rows of column names that need to be replaced with one row of names. dominoSignal takes TF regulons as a named list of target-gene vectors, so regulons from any TF activity method can be used. We provide a helper function, `create_regulon_list_scenic()` to create this list from the output of the [pySCENIC](https://pyscenic.readthedocs.io/en/latest/) `ctx` function.

```{r Create Regulon List}
regulons <- regulons[-1:-2, ]
colnames(regulons) <- c(
    "TF", "MotifID", "AUC", "NES", "MotifSimilarityQvalue", "OrthologousIdentity",
    "Annotation", "Context", "TargetGenes", "RankAtMax"
)
regulon_list <- create_regulon_list_scenic(regulons = regulons)
```

Users should be aware that the area under the recovery curve (AUC) matrix from [SCENIC](https://scenic.aertslab.org/) is loaded in a cell x TF orientation and should be transposed to TF x cell orientation. pySCENIC also appends "(+)" to TF names that are converted to "..." upon loading into R. These characters can be included without affecting the results of dominoSignal analysis but can be confusing when querying TF features in the data. We recommend comprehensive removal of the "..." characters using the `gsub()` function.

```{r Orient AUC Matrix}
auc_in <- as.data.frame(t(auc))
# Remove pattern "..." from the end of all rownames:
rownames(auc_in) <- gsub("\\.\\.\\.$", "", rownames(auc_in))
```

### Loading the CellPhoneDB database

dominoSignal reads receptor-ligand databases in a uniform data frame format referred to as a receptor-ligand map (`rl_map`) so any reference database can be used if formatted appropriately. Each row corresponds to a receptor-ligand interaction. Genes participating in an interaction are referred to as "partner A" and "partner B" without requiring for fixed ordering of whether A or B is the ligand and vice versa. The minimum required columns of this data frame are:

- `int_pair`: the names of the interacting ligand and receptor separated by " & "
- `gene_A`: the gene or genes encoding partner A
- `gene_B`: the gene or genes encoding partner B
- `type_A`: (`"L"`, `"R"`) - indicates whether partner A is a ligand (`"L"`) or receptor (`"R"`)
- `type_B`: (`"L"`, `"R"`) - indicates whether partner B is a ligand (`"L"`) or receptor (`"R"`)

Additional annotation columns can be provided such as `name_A` and `name_B` for ligands or receptors whose name in the interaction database does not match the names of their encoding genes. This formatting also allows for the consideration of ligand and receptor complexes comprised of a heteromeric combination of multiple proteins that must be co-expressed to function. In these cases, the `name_*` column shows the name of the protein complex, and the `gene_*` column shows the names of the genes encoding the components of the complex separated by commas. When plotting results from the build `domino()` object, the names of the interacting ligands and receptors will be used based on combined expression of the complex components.

To facilitate the use of this formatting with the CellPhoneDB database, we include a helper function, `create_rl_map_cellphonedb()`, that parses files from the CellPhoneDB database to arrive at the `rl_map` format.

```{r Load CellPhoneDB Database}
complexes <- read.csv(
    paste0(
        cellphone_data, "/complex_input.csv"
    ),
    stringsAsFactors = FALSE
)
genes <- read.csv(
    paste0(
        cellphone_data, "/gene_input.csv"
    ),
    stringsAsFactors = FALSE
)
interactions <- read.csv(
    paste0(
        cellphone_data, "/interaction_input.csv"
    ),
    stringsAsFactors = FALSE
)
proteins <- read.csv(
    paste0(
        cellphone_data, "/protein_input.csv"
    ),
    stringsAsFactors = FALSE
)

rl_map <- create_rl_map_cellphonedb(
    genes = genes,
    proteins = proteins,
    interactions = interactions,
    complexes = complexes,
    database_name = "CellPhoneDB_v4.0" # database version used
)
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(head(rl_map),
    caption = "Receptor-ligand map of interactions."),
    bootstrap_options = c("striped", "hover"))
```

</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(head(rl_map),
    caption = "Receptor-ligand map of interactions."),
    bootstrap_options = c("striped", "hover"))
```

#### Optional: Adding interactions manually
Interactions missing from the database can be appended manually by formatting them as a data frame with the same column headers as the `rl_map` and using the `rbind()` function.

```{r Appending interactions}
# Integrin complexes are not annotated as receptors in CellPhoneDB_v4.0.
# Collagen-integrin interactions between cells may be missed unless tables
# from the CellPhoneDB reference are edited or the interactions are manually added.

col_int_df <- data.frame(
    "int_pair" = "a11b1 complex & COLA1_HUMAN",
    "name_A" = "a11b1 complex", "uniprot_A" = "P05556,Q9UKX5", "gene_A" = "ITGB1,ITGA11", "type_A" = "R",
    "name_B" = "COLA1_HUMAN", "uniprot_B" = "P02452,P08123", "gene_B" = "COL1A1,COL1A2", "type_B" = "L",
    "annotation_strategy" = "manual", "source" = "manual", "database_name" = "manual"
)
rl_map_append <- rbind(col_int_df, rl_map)
```

<details><summary>Show table code</summary>

```{r, eval = FALSE}
kableExtra::kable_styling(knitr::kable(head(rl_map_append),
    caption = "Manual rl_map of collagen-integrin interactions."),
    bootstrap_options = c("striped", "hover"))
```

</details>

```{r, echo = FALSE}
kableExtra::kable_styling(knitr::kable(head(rl_map_append),
    caption = "Manual rl_map of collagen-integrin interactions."),
    bootstrap_options = c("striped", "hover"))
```

## Analysis with domino objects

dominoSignal analysis takes place in two steps. 

1. `create_domino()`

    a. Initializes the `domino()` result object

    b. Assesses differential TF activity across cell clusters by Wilcoxon rank-sum test (when `tf_selection_method = "clusters"`, the default)

    c. Establishes TF-receptor linkages based on Spearman correlation of TF activities with receptor expression across the queried data set. 

2. `build_domino()`

    a. Sets the parameters for which TFs and receptors are called as active within a cell cluster

    b. Aggregates the scaled expression of ligands capable of interacting with the active receptors for assessment of ligand type and cellular source triggering activation of a receptor.

### Required inputs from data set

dominoSignal infers active receipt of signals via receptors based on the correlation of the receptor's expression with TF activity across the data set and differential activity of the TF within a cell cluster. Correlations are conducted using scaled expression values rather than raw counts or normalized counts. A named (factor) vector of cluster assignments, where the names are cell barcodes matching the expression matrix, is required. Assessing signaling based on other categorical groupings of cells can be achieved by passing these groupings as `clusters` to `build_domino()` in place of cell type cluster labels. dominoSignal accepts a matrix of counts, a matrix of scaled counts, and a **named** vector of cell cluster labels **as factors**. Shown below is how to extract these elements from a [SingleCellExperiment](https://doi.org/doi:10.18129/B9.bioc.SingleCellExperiment) object. Note that since the data is scaled by gene, genes with no expression in any cell need to be removed.

```{r load cell features}
counts <- assay(pbmc, "counts")
logcounts <- assay(pbmc, "logcounts")
logcounts <- logcounts[rowSums(logcounts) > 0, ]
z_scores <- t(scale(t(logcounts)))
clusters <- factor(pbmc$cell_type)
names(clusters) <- colnames(pbmc)
```

Note: Ligand and receptor expression can only be assessed for genes included in the z\_scores matrix. Many scRNAseq analysis pipelines recommend storing only genes with high variance in scaled expression slots for these data objects, thereby missing many genes encoding ligands and receptors. **Ensure that all your genes of interest are included in the rows of your `z_scores` matrix.** Scaled expression was calculated for all genes in this PBMC data set after removal of genes expressed in less than three cells. 

### Create domino object

At this point, the `create_domino()` function can be used to make the object. Parameters of note include `use_complexes` (which determines whether heteromeric complexes will be considered in the analysis; if TRUE, a complex is only linked to its partner if the median correlation of its component genes exceeds the threshold), and `remove_rec_dropout` (which decides whether receptor values of zero will be considered in correlation calculations; this measure is intended to reduce the effect of dropout on receptors with low expression but should be used with caution as there is no way to distinguish between technical dropout and biological absence of expression).

To run `create_domino()` with matrix and vector inputs:
```{r Create Domino, results='hide', warning=FALSE, tidy = FALSE}
pbmc_dom <- create_domino(
    rl_map = rl_map, # receptor-ligand map data frame
    features = auc_in, # TF activity scores (AUC matrix)
    counts = counts, # counts matrix
    z_scores = z_scores, # scaled expression matrix
    clusters = clusters, # factored, named vector of cluster assignments
    tf_targets = regulon_list, # list of TFs and their regulons
    use_complexes = TRUE, # include interactions with complexes?
    remove_rec_dropout = FALSE # remove zeroes from correlation calculations?
)
```

The parameters used to create the object are stored in the `misc` slot of the `domino()` object. For information on what is stored in a `domino()` object and how to access it, please see our [vignette on the structure of domino objects](https://fertiglab.github.io/dominoSignal/articles/domino_object_vignette).

### Build the domino network

`build_domino()` finalizes the construction of the `domino()` object by setting parameters for identifying active TFs, receptor linkage with TFs based on magnitude of positive correlation, and the minimum percentage of cells within a cluster that have expression of a receptor for the receptor to be called as active.

There are also options for thresholds of the number of TFs that may be called active in a cluster and the number of receptors that may be linked to any one TF. For thresholds of _n_ TFs and _m_ receptors, the bottom _n_ TFs by lowest p-values from the Wilcoxon rank-sum test and the top _m_ receptors by Spearman correlation coefficient are chosen.

```{r Build Domino, tidy = FALSE}
pbmc_dom <- build_domino(
    dom = pbmc_dom,
    max_tf_pval = 0.001, # threshold for p-value of DE for TFs
    max_tf_per_clust = 25, # maximum number of TFs to call active in cluster
    max_rec_per_tf = 25, # maximum number receptors to link to a TF
    rec_tf_cor_threshold = 0.25, # minimum correlation between receptor and TF
    min_rec_percentage = 0.1 # minimum % of cells in cluster with expression to include receptor
)
```

Thresholds for both the number of receptors and TFs can be sent to infinity (Inf) to collect all receptors and TFs that meet other thresholds.

```{r Build Domino All Significant Interactions, tidy = TRUE, eval=FALSE}
pbmc_dom_all <- build_domino(
    dom = pbmc_dom,
    max_tf_pval = 0.001,
    max_tf_per_clust = Inf,
    max_rec_per_tf = Inf,
    rec_tf_cor_threshold = 0.25,
    min_rec_percentage = 0.1
)
```

If the `tf_selection_method` argument is set to "variable" or "all" in `create_domino()`, the `max_tf_pval` argument is ignored. The selected TFs are ranked by their maximum correlation with expressed receptors in the cluster (those that mean the `min_rec_percentage` threshold), subsetting to only include those with a correlation above `rec_tf_cor_threshold`. The top `max_tf_per_clust` TFs are then included in linkages for the cluster.

```{r create build domino all or variable, tidy = TRUE, eval = FALSE}
pbmc_create_variable <- create_domino(
    rl_map = rl_map, # receptor-ligand map data frame
    features = auc_in, # TF activity scores (AUC matrix)
    counts = counts, # counts matrix
    z_scores = z_scores, # scaled expression matrix
    clusters = clusters, # factored, named vector of cluster assignments
    tf_targets = regulon_list, # list of TFs and their regulons
    use_complexes = TRUE, # include interactions with complexes?
    remove_rec_dropout = FALSE, # remove zeroes from correlation calculations?
    tf_selection_method = "variable", # select TFs based on variation
    tf_variance_quantile = 0.25 # select TFs with variance above this quantile
)

pbmc_variable <- build_domino(
    dom = pbmc_create_variable,
    max_tf_pval = 0.001, # ignored if tf_selection_method isn't "clusters"
    max_tf_per_clust = 25, # maximum number of TFs to call active in cluster
    max_rec_per_tf = 25, # maximum number receptors to link to a TF
    rec_tf_cor_threshold = 0.25, # minimum correlation between receptor and TF
    min_rec_percentage = 0.1 # minimum % of cells in cluster with expression to include receptor
)
```

## Visualization of domino results

Multiple functions are available to visualize the intracellular networks between receptors and TFs and the receptor-ligand mediated intercellular networks between clusters.

### Summarize TF activity and linkages

Enrichment of TF activity by cluster can be visualized by `feat_heatmap()` which plots data set-wide TF activity scores as a heatmap.

```{r}
#| fig.alt: >
#|  Heatmap showing feature activity for cells ordered by cluster.
feat_heatmap(pbmc_dom,
    norm = TRUE, bool = FALSE, use_raster = FALSE,
    row_names_gp = grid::gpar(fontsize = 4)
)
```

### Cumulative signaling between clusters

The cumulative degree of signaling between clusters is assessed as the sum of the mean z-scored expression of ligands targeting active receptors on another cluster. This can be visualized in graph format using the `signaling_network()` function. Nodes represent each cell cluster and edges scale with the magnitude of signaling between the clusters. The color of the edge corresponds to the sender cluster for that signal.

```{r}
#| fig.alt: >
#|  Network plot where nodes are clusters and edges indicate the magnitude of
#|  communication between the clusters.
signaling_network(pbmc_dom, edge_weight = 0.5, max_thresh = 3)
```

Signaling networks can also be drawn with the edges only rendering the signals directed towards a given cell type or signals from one cell type directed to others. To see those options in use, as well as other plotting functions and their options, please see our [plotting vignette](https://fertiglab.github.io/dominoSignal/articles/plotting_vignette.html).

### Specific Signaling Interactions between Clusters

Beyond the aggregated degree of signaling between clusters, the degrees of signaling through specific receptor-ligand interactions can be assessed. `gene_network()` provides a graph of linkages between active TFs in a cluster, the linked receptors in that cluster, and the possible ligands of these active receptors.

```{r}
#| fig.alt: >
#|  Network plot with columns corresponding to ligands, receptors, and transcription factors
#|  for signaling to dendritic cells.

gene_network(pbmc_dom, clust = "dendritic_cell", layout = "grid")
```

New to dominoSignal, `gene_network()` can be used between two clusters to determine if any of the possible ligands for a given receptor are expressed by a putative outgoing signaling cluster.

```{r}
#| fig.alt: >
#|  Network plot with columns corresponding to ligands, receptors, and transcription factors
#|  showing signaling to dendritic cells from the CD14 monocyte population.
gene_network(pbmc_dom,
    clust = "dendritic_cell", OutgoingSignalingClust = "CD14_monocyte",
    layout = "grid"
)
```

A comprehensive assessment of ligand expression targeting active receptors on a given cluster can be assessed with `incoming_signaling_heatmap()`.

```{r, out.width = "95%"}
#| fig.alt: >
#|  Heatmap showing ligand expression from each sending population to receiving receptors in
#|  dendritic cell population
incoming_signaling_heatmap(pbmc_dom, rec_clust = "dendritic_cell", max_thresh = 2.5, use_raster = FALSE)
```

Another form of comprehensive ligand expression assessment is available for individual active receptors in the form of circos plots. The outer arcs correspond to clusters in the `domino()` object with inner arcs representing each possible ligand of the plotted receptor. Arcs are drawn between ligands on a cell type and the receptor if the ligand is expressed above the specified threshold. Arc widths correspond to the mean expression of the ligand by the cluster with the widest arc width scaling to the maximum expression of the ligand within the data.

```{r, fig.asp = 0.6, out.width = "90%"}
#| fig.alt: >
#|  A circos plot with outer arcs corresponding to clusters and inner arcs corresponding to ligands
#|  directed to receptor CD74. Arcs represent mean expression of the ligands.
circos_ligand_receptor(pbmc_dom, receptor = "CD74")
```

## Example run

Here is an example of the complete dominoSignal workflow with default settings using the PBMC data set. We start from the `pbmc` SingleCellExperiment object, the SCENIC results, and the CellPhoneDB files. In this example, we focus on signaling to the CD14 monocyte cluster.

```{r eval = FALSE}
# Read and format SCENIC results
regulons <- read.csv(scenic_regulon)
auc <- read.table(scenic_auc,
    header = TRUE, row.names = 1,
    stringsAsFactors = FALSE, sep = ","
)
regulons <- regulons[-1:-2, ]
colnames(regulons) <- c(
    "TF", "MotifID", "AUC", "NES", "MotifSimilarityQvalue", "OrthologousIdentity",
    "Annotation", "Context", "TargetGenes", "RankAtMax"
)
regulon_list <- create_regulon_list_scenic(regulons = regulons)
auc_in <- as.data.frame(t(auc))
rownames(auc_in) <- gsub("\\.\\.\\.$", "", rownames(auc_in))

# Create rl_map
complexes <- read.csv(paste0(cellphone_data, "/complex_input.csv"), stringsAsFactors = FALSE)
genes <- read.csv(paste0(cellphone_data, "/gene_input.csv"), stringsAsFactors = FALSE)
interactions <- read.csv(paste0(cellphone_data, "/interaction_input.csv"), stringsAsFactors = FALSE)
proteins <- read.csv(paste0(cellphone_data, "/protein_input.csv"), stringsAsFactors = FALSE)

rl_map <- create_rl_map_cellphonedb(
    genes = genes,
    proteins = proteins,
    interactions = interactions,
    complexes = complexes,
    database_name = "CellPhoneDB_v4.0" # database version used
)

# Get inputs from SingleCellExperiment object
counts <- assay(pbmc, "counts")
logcounts <- assay(pbmc, "logcounts")
logcounts <- logcounts[rowSums(logcounts) > 0, ]
z_scores <- t(scale(t(logcounts)))
clusters <- factor(pbmc$cell_type)
names(clusters) <- colnames(pbmc)

# Infer network
pbmc_dom <- create_domino(rl_map = rl_map, features = auc_in, counts = counts,
    z_scores = z_scores, clusters = clusters, tf_targets = regulon_list,
    use_complexes = TRUE, remove_rec_dropout = FALSE)

pbmc_dom <- build_domino(dom = pbmc_dom)

# Visualize results
feat_heatmap(pbmc_dom)
signaling_network(pbmc_dom, showIncomingSignalingClusts = "CD14_monocyte")
gene_network(pbmc_dom, clust = "CD14_monocyte")
incoming_signaling_heatmap(pbmc_dom, rec_clust = "CD14_monocyte")
circos_ligand_receptor(pbmc_dom, receptor = "CD74")
```

# Related articles

- [Interacting with domino objects](https://fertiglab.github.io/dominoSignal/articles/domino_object_vignette.html): object structure and accessor functions
- [Plotting functions and options](https://fertiglab.github.io/dominoSignal/articles/plotting_vignette.html): visualization options
- [Differential signaling workflow](https://fertiglab.github.io/dominoSignal/articles/differential_signaling.html): comparing signaling across subjects
- [SCENIC for TF activity scoring](https://fertiglab.github.io/dominoSignal/articles/tf_scenic_vignette.html): generating TF activity inputs
- [Using the CellPhoneDB database](https://fertiglab.github.io/dominoSignal/articles/cellphonedb_vignette.html): preparing a receptor-ligand map

***

To report a bug, ask a question, or suggest a feature, please [open an issue](https://github.com/FertigLab/dominoSignal/issues).

<details><summary>Vignette Build Information</summary>
Date last built and session information:
```{r}
Sys.Date()
sessionInfo()
```
</details>

