---
title: "Interacting with domino Objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interacting with domino Objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.cap = "",
  tidy = TRUE
)
options(timeout = 600)
```


In this tutorial, we will go into more detail on the structure of a `domino()` object and the ways in which to access the data stored within. We will be using the `domino()` object we built on the [Getting Started](https://fertiglab.github.io/dominoSignal) page. If you have not yet built a `domino()` object, you can do so by following the instructions on that page.

<details><summary>Instructions to load data</summary>

```{r setup}
set.seed(42)
library(dominoSignal)

# BiocFileCache helps with file management across sessions
bfc <- BiocFileCache::BiocFileCache(ask = FALSE)
data_url <- "https://zenodo.org/records/10951634/files/pbmc_domino_built.rds"
tmp_path <- BiocFileCache::bfcrpath(bfc, data_url)
dom <- readRDS(tmp_path)
```

</details>

## Object contents

There is a great deal of information stored with the `domino()` object class. The `domino()` object is an S4 class object that contains a variety of information about the data set used to build the object, the calculated values, and the linkages between receptors, ligands, and transcription factors. The object is structured as follows (with some examples of the information stored within each slot):

- Input Data

  - Information about the database used to construct the `rl_map`

  - Inputted counts matrix
  
  - Inputted z-scored counts matrix
  
  - Inputted cluster labels
  
  - Inputted transcription factor (TF) activity scores

- Calculated values

  - P-values for differential activity of TFs in each cluster

  - Correlation values between receptor expression and TF activity

  - Receptor-TF correlations with receptor complexes summarized as the median of their component genes

- Linkages

  - Complexes show the component genes of any complexes in the `rl_map`

  - Receptor-ligand linkages as determined from the `rl_map`

  - TF-target linkages as determined from the SCENIC analysis (or other regulon inference method)

  - TFs that are active in each cluster

  - TFs that are correlated with receptors

  - TFs that are correlated with receptors in each cluster

  - Receptors which are active in each cluster

  - Ligands that may activate a receptor in a given cluster (so-called incoming ligands; these may include ligands from outside the data set)

- Signaling matrices

  - For each cluster, incoming ligands and the clusters within the data set that they are coming from

  - A summary of signaling between all clusters

- Miscellaneous Information

  - Create information, which includes the parameters and package version used to create the object in `create_domino()`

  - Build information, which includes the parameters and package versionused to build the object in `build_domino()`

  - The processed `rl_map` used in building the object

  - The percent expression of receptors within each cluster

For commonly accessed information (the number of cells, clusters, and some build information), the show and print methods for `domino()` objects can be used.

```{r show, message = TRUE}
dom
```

```{r print, message = TRUE}
print(dom)
```

## Access functions

To facilitate access to the information stored in the `domino()` object, we have provided a collection of functions to retrieve specific items. These functions begin with "dom_" and can be listed using `ls()`.

```{r list-functions}
ls("package:dominoSignal", pattern = "^dom_")
```

### Input data

When creating a `domino()` object with the `create_domino()` function, several inputs are required which are then stored in the `domino()` object itself. These include cluster labels, the counts matrix, z-scored expression matrix, TF activity scores, and the receptor-ligand database used in `create_rl_map_cellphonedb()`.

For example, to access the cluster names in the `domino()` object:

```{r cluster-names}
dom_clusters(dom)
```

Setting an argument `labels = TRUE` will return the vector of cluster labels for each cell rather than the unique cluster names.

To access the counts:

```{r counts}
count_matrix <- dom_counts(dom)
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(count_matrix[1:5, 1:5], caption = "Subset of counts matrix from domino object."),
    bootstrap_options = c("striped", "hover"))
```
</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(count_matrix[1:5, 1:5], caption = "Subset of counts matrix from domino object."),
    bootstrap_options = c("striped", "hover"))
```

Or z-scored expression:

```{r zcounts}
z_matrix <- dom_zscores(dom)
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(z_matrix[1:5, 1:5], caption = "Subset of z-scored counts matrix from domino object."),
    bootstrap_options = c("striped", "hover"))
```
</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(z_matrix[1:5, 1:5], caption = "Subset of z-scored counts matrix from domino object."),
    bootstrap_options = c("striped", "hover"))
```

The TF activity scores can be similarly accessed:

```{r tf_activity}
activity_matrix <- dom_tf_activation(dom)
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(activity_matrix[1:5, 1:5], caption = "Subset of transcription factor activity matrix from domino object."),
    bootstrap_options = c("striped", "hover"))
```

</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(activity_matrix[1:5, 1:5], caption = "Subset of transcription factor activity matrix from domino object."),
    bootstrap_options = c("striped", "hover"))
```

Information about the database referenced for receptor-ligand pairs and composition of protein complexes can be extracted from the `dom_database()` function. By default, the function returns the name(s) of the database(s) used:

```{r db-name}
dom_database(dom)
```

If you would like to view the entire receptor-ligand map, set `name_only = FALSE`:

```{r db-all}
db_matrix <- dom_database(dom, name_only = FALSE)
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(db_matrix[1:5, 1:5],
    caption = "Subset of full receptor-ligand map from domino object."),
    bootstrap_options = c("striped", "hover"))
```

</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(db_matrix[1:5, 1:5],
    caption = "Subset of full receptor-ligand map from domino object."),
    bootstrap_options = c("striped", "hover"))
```

### Calculations

By default, active TFs in each cluster are determined by conducting Wilcoxon rank-sum tests for each TF where the TF activity scores amongst all cells in a cluster are tested against the activity scores of all cells outside of the cluster. The p-values for the one-sided test for greater activity within the cluster compared to other cells can be accessed with the `dom_de()` function.

```{r de}
de_matrix <- dom_de(dom)
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(de_matrix[1:5, 1:5],
    caption = "Subset of differential expression p-values for TFs."),
    bootstrap_options = c("striped", "hover"))
```

</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(de_matrix[1:5, 1:5],
    caption = "Subset of differential expression p-values for TFs."),
    bootstrap_options = c("striped", "hover"))
```

Linkages between receptors and TFs are assessed by Spearman correlation between TF activity scores and scaled expression of receptor-encoding genes across all cells in the data set. Spearman coefficients can be accessed with the `dom_correlations()` function. The default (`type = "rl"`) will return the correlation of each receptor gene with each TF. Setting `type= "complex"` will return receptor-TF correlations in which receptor complexes are summarized as the median of their component genes (the values used by `build_domino()`).

```{r correlations}
cor_matrix <- dom_correlations(dom)
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(cor_matrix[1:5, 1:5],
    caption = "Correlation matrix from domino object."),
    bootstrap_options = c("striped", "hover"))
```

</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(cor_matrix[1:5, 1:5],
    caption = "Correlation matrix from domino object."),
    bootstrap_options = c("striped", "hover"))
```

### Linkages

Linkages between ligands, receptors, and TFs can be accessed in several different ways, depending on the specific link and the scope desired. The `dom_linkages()` function has three arguments - the first, like all of our access functions, is for the `domino()` object. The second, `link_type`, is used to specify which linkages are desired (options are complexes, receptor-ligand, tf-target, or tf-receptor). The third argument, `by_cluster`, determines whether the linkages returned are arranged by cluster (though this does change the available linkage types to tf-receptor, receptor, or incoming-ligand). For example, to access the complexes used across the dataset:

```{r, complex}
complex_links <- dom_linkages(dom, link_type = "complexes")
# Look for components of IL7 receptor complex
complex_links$IL7_receptor
```

To view incoming ligands to each cluster:

```{r, lig-by-clust}
incoming_links <- dom_linkages(dom, link_type = "incoming-ligand", by_cluster = TRUE)
# Check incoming signals to dendritic cells
incoming_links$dendritic_cell
```

If, for some reason, you find yourself in need of the entire linkage structure (not recommended), it can be accessed through its slot name; `domino()` objects are S4 objects.

```{r link}
all_linkages <- slot(dom, "linkages")
# Names of all sub-structures:
names(all_linkages)
```

Alternately, to obtain a simplified list of receptors, ligands, and\/or features in the `domino()` object, use the `dom_network_items()` function. To pull all transcription factors associated with the dendritic cell cluster:

```{r collate}
dc_tfs <- dom_network_items(dom, "dendritic_cell", which_return = "features")
head(dc_tfs)
```

### Signaling Matrices

The mean z-scored expression of ligands, by sending cluster, that target each receiving cluster's active receptors can be accessed in matrix form.

```{r global-signalling}
signal_matrix <- dom_signaling(dom)
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(signal_matrix,
    caption = "Cluster signaling matrix of averaged z_scored expression."),
    bootstrap_options = c("striped", "hover"))
```

</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(signal_matrix,
    caption = "Cluster signaling matrix of averaged z_scored expression."),
    bootstrap_options = c("striped", "hover"))
```


To view signaling to a specific cluster from the other clusters, set the `cluster` argument to the cluster name.

```{r clust-signal}
dc_matrix <- dom_signaling(dom, "dendritic_cell")
```

<details><summary>Show table code</summary>

```{r eval = FALSE}
kableExtra::kable_styling(knitr::kable(dc_matrix,
    caption = "Cluster signaling matrix for dendritic cell cluster."),
    bootstrap_options = c("striped", "hover"))
```

</details>

```{r echo = FALSE}
kableExtra::kable_styling(knitr::kable(dc_matrix,
    caption = "Cluster signaling matrix for dendritic cell cluster."),
    bootstrap_options = c("striped", "hover"))
```

### Build information

To keep track of the options set when running `create_domino()` and `build_domino()`, they are stored within the `domino()` object itself. To view these options, use the `dom_info()` function. Objects created with dominoSignal version 1.7.1 or newer also record the package versions used to create and build them. This example object was generated before that, so the package versions are `NULL` (as are the `build_domino()` parameters, which were also added in version 1.7.1).

```{r info}
dom_info(dom)
```

Building the object again (with different parameters) results in a `domino()` object with the new parameters.

```{r rebuild}
dom <- build_domino(dom, max_tf_per_clust = Inf, max_tf_pval = 0.05,
    max_rec_per_tf = 50, rec_tf_cor_threshold = 0.1, min_rec_percentage = 0.05)
dom_info(dom)
```

## Related articles

- [Get started with dominoSignal](https://fertiglab.github.io/dominoSignal/articles/dominoSignal.html): the basic workflow
- [Plotting functions and options](https://fertiglab.github.io/dominoSignal/articles/plotting_vignette.html): visualization options
- [Differential signaling workflow](https://fertiglab.github.io/dominoSignal/articles/differential_signaling.html): comparing signaling across subjects
- [SCENIC for TF activity scoring](https://fertiglab.github.io/dominoSignal/articles/tf_scenic_vignette.html): generating TF activity inputs
- [Using the CellPhoneDB database](https://fertiglab.github.io/dominoSignal/articles/cellphonedb_vignette.html): preparing a receptor-ligand map

***

To report a bug, ask a question, or suggest a feature, please [open an issue](https://github.com/FertigLab/dominoSignal/issues).

<details><summary>Vignette Build Information</summary>
Date last built and session information:
```{r}
Sys.Date()
sessionInfo()
```

</details>
