---
title: "Using the CellPhoneDB Database"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.cap = "",
    tidy = TRUE
)
options(timeout = 300)
```

The dominoSignal analysis pipeline requires the use of a reference receptor-ligand database to identify potential receptor-ligand interactions. We recommend CellPhoneDB. Here is a very brief tutorial on how to download the requisite files from CellPhoneDB v4.0.0.

## File downloads for CellPhoneDB:

Database files from [CellPhoneDB v4.0.0](https://github.com/ventolab/cellphonedb-data/releases/tag/v4.0.0) for human scRNAseq data can be installed from a public GitHub repository from the Teichmann Group, which developed CellPhoneDB.

```{r download CellPhoneDB, eval = FALSE}
# directories for the download
temp_dir <- tempdir()
cellphone_dir <- file.path(temp_dir, "cellphone")

# URL for desired version of CellPhoneDB
cellphone_url <- "https://github.com/ventolab/cellphonedb-data/archive/refs/tags/v4.0.0.tar.gz"`

# download compressed database
cellphone_tar <- paste0(temp_dir, "/cellphoneDB_v4.tar.gz")
download.file(url = cellphone_url, destfile = cellphone_tar)

# move contents of the compressed file to a new directory
untar(tarfile = cellphone_tar, exdir = cellphone_dir)
cellphone_data <- paste0(cellphone_dir, "/cellphonedb-data-4.0.0/data")
```

To facilitate the use of these files in the format used in dominoSignal, we include a helper function, `create_rl_map_cellphonedb()`, that automatically parses files from the CellPhoneDB database to arrive at the rl\_map format.

## Related articles

- [Get started with dominoSignal](https://fertiglab.github.io/dominoSignal/articles/dominoSignal.html): the basic workflow
- [Interacting with domino objects](https://fertiglab.github.io/dominoSignal/articles/domino_object_vignette.html): object structure and accessor functions
- [Plotting functions and options](https://fertiglab.github.io/dominoSignal/articles/plotting_vignette.html): visualization options
- [Differential signaling workflow](https://fertiglab.github.io/dominoSignal/articles/differential_signaling.html): comparing signaling across subjects
- [SCENIC for TF activity scoring](https://fertiglab.github.io/dominoSignal/articles/tf_scenic_vignette.html): generating TF activity inputs

***

To report a bug, ask a question, or suggest a feature, please [open an issue](https://github.com/FertigLab/dominoSignal/issues).

<details><summary>Vignette Build Information</summary>
Date last built and session information:
```{r}
Sys.Date()
sessionInfo()
```
</details>