---
title: "SCENIC for TF Activity Scoring"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.cap = "",
    tidy = TRUE
)
options(timeout = 300)
```

[SCENIC](https://scenic.aertslab.org/) is our preferred software for TF activity scoring. We recommend using the Python implementation ([pySCENIC](https://pyscenic.readthedocs.io/en/latest/)) as it is faster than the original R implementation. Be aware, the use of [SCENIC](https://scenic.aertslab.org/) is often the slowest and most memory intensive step of this analysis pipeline. [SCENIC](https://scenic.aertslab.org/) should be run on computing resources with access to multi-core processing and large amounts of memory.

## File downloads for SCENIC
For this tutorial, [pySCENIC](https://pyscenic.readthedocs.io/en/latest/) is used as the method for TF activity inference. This requires downloading some files prior to use. Here, we show how to download these files and the way we run pySCENIC to generate the necessary files for use in dominoSignal analysis.

A singularity image of [SCENIC](https://scenic.aertslab.org/) v0.12.1 can be built from the aertslab Docker Hub image. [SCENIC](https://scenic.aertslab.org/) requires a list of TFs, motif annotations, and cisTarget motifs which are all available from the authors of SCENIC for human (HGNC), mouse (MGI), and fly. The following will download everything necessary for an analysis of a data set with HGNC gene labels for the hg38 genome.


### Reference files

We build the singularity image and download reference files to the temporary directory.

```{bash, eval=FALSE}
SCENIC_DIR="${TMPDIR:-/tmp}/scenic"
mkdir ${SCENIC_DIR}

# Build singularity image
singularity build "${SCENIC_DIR}/aertslab-pyscenic-0.12.1.sif" docker://aertslab/pyscenic:0.12.1

# Matrix containing motifs as rows and genes as columns and ranking position for each gene and motif (based on CRM scores) as values. URLs provided link to v2 feather files required for the 0.12.1 version of pySENIC.

curl "https://resources.aertslab.org/cistarget/databases/homo_sapiens/hg38/refseq_r80/mc_v10_clust/gene_based/hg38_10kbp_up_10kbp_down_full_tx_v10_clust.genes_vs_motifs.rankings.feather" \
  -o "${SCENIC_DIR}/hg38_10kbp_up_10kbp_down_full_tx_v10_clust.genes_vs_motifs.rankings.feather"

curl "https://resources.aertslab.org/cistarget/databases/homo_sapiens/hg38/refseq_r80/mc_v10_clust/gene_based/hg38_500bp_up_100bp_down_full_tx_v10_clust.genes_vs_motifs.rankings.feather" \
  -o "${SCENIC_DIR}/hg38_500bp_up_100bp_down_full_tx_v10_clust.genes_vs_motifs.rankings.feather"

# List of genes encoding TFs in the hg38 reference genome
curl "https://resources.aertslab.org/cistarget/tf_lists/allTFs_hg38.txt" \
  -o "${SCENIC_DIR}/allTFs_hg38.txt"

# Motif annotations based on the 2017 cisTarget motif collection. Use these files if you are using the mc9nr databases.
curl "https://resources.aertslab.org/cistarget/motif2tf/motifs-v9-nr.hgnc-m0.001-o0.0.tbl" \
  -o "${SCENIC_DIR}/motifs-v9-nr.hgnc-m0.001-o0.0.tbl"
```

### Preparing preprocessed data:
These examples use the `pbmc` SingleCellExperiment object and the `input_dir` directory created in the [Get Started](https://fertiglab.github.io/dominoSignal/articles/dominoSignal.html) vignette.

dominoSignal was designed to be compatible with single cell workflows, accepting gene by cell matrices for counts and scaled counts. However, [pySCENIC](https://pyscenic.readthedocs.io/en/latest/) is used for TF activity inference, and it requires as input a cell by gene matrix. This is the opposite orientation of many R based single cell packages (such as `{Seurat}` or [SingleCellExperiment](https://doi.org/doi:10.18129/B9.bioc.SingleCellExperiment)) but is the default for Python based tools such as [scanpy](https://scanpy.readthedocs.io/en/stable/). The RNA counts matrix can be saved as a tab-separated value (.tsv) or comma-separated value (.csv) file after transposing the matrix to cell by gene orientation. Below is an example of extracting the counts from a `SingleCellExperiment` object and saving it as a .tsv file.

```{r Save Counts Matrix as tsv File, eval=FALSE}
pbmc_counts <- assay(pbmc, "counts")
write.table(t(as.matrix(pbmc_counts)), paste0(input_dir, "/pbmc3k_counts.tsv"),
    sep = "\t", col.names = NA
)
```

TSV and CSV files are very inefficient for storing large matrices. As an alternative, the counts matrix can be saved as a loom file and directly passed to [pySCENIC](https://pyscenic.readthedocs.io/en/latest/) in this format. Generating a loom file in R requires use of the [loomR](https://github.com/mojaveazure/loomR) package. The package automatically handles the transposition of the counts matrix to cell by gene orientation as well. We recommend this approach to passing a counts matrix to pySCENIC. However, users should be aware that `loomR` is not hosted on CRAN or Bioconductor at the time of this vignette's creation. 

```{r Save Counts as loomR, eval = FALSE}
library(loomR)

# save loom counts matrix
pbmc_counts <- assay(pbmc, "counts")
pbmc_loom <- loomR::create(
    filename = paste0(input_dir, "/pbmc3k_counts.loom"),
    data = pbmc_counts
)
# Depending on the version of loomR, you may need to manually add CellID and Gene attributes:
pbmc_loom$add.row.attribute(list(Gene = rownames(pbmc)))
pbmc_loom$add.col.attribute(list(CellID = colnames(pbmc)))

# connection to the loom file must be closed to complete file creation
pbmc_loom$close_all()
```

We will use the `pbmc3k\_counts.loom` file in this tutorial. Copy or link it into `${SCENIC_DIR}` for the following steps.

## Running SCENIC

[pySCENIC](https://pyscenic.readthedocs.io/en/latest/) is initiated using bash scripting in the terminal. The analysis consists of three steps to score genes for TF motif enrichment, construct TF regulons consisting of genes targeted by the TFs, and arrive at AUC scores for enrichment of regulon gene transcription within each cell.

### grn: construct TF-modules

Co-expression modules are used to quantify gene-TF adjacencies. Arguments are collected in a bash array so they can be annotated and the command still runs as written (this is not necessary for analysis; arguments can be passed directly to the command).


```{bash GRN, eval=FALSE, tidy = FALSE}
grn_args=(
    "${SCENIC_DIR}/pbmc3k_counts.loom"   # path to loom file
    "${SCENIC_DIR}/allTFs_hg38.txt"      # list of TFs
    -o "${SCENIC_DIR}/pbmc_adj_3k.tsv"   # output adjacency table
    --num_workers 6                      # number of CPUs if multi-core processing is available
    --seed 123                           # specify a random seed for reproducibility
)
singularity exec "${SCENIC_DIR}/aertslab-pyscenic-0.12.1.sif" pyscenic grn "${grn_args[@]}"
```

### ctx: construct TF regulons with pruning based on TF motif enrichment

The rankings of genes based on enrichment of TF motifs to the transcription start site (TSS) are considered in the construction of regulons, where target genes in the TF-modules are removed if they lack motifs proximal to the TSS where the TF may bind.

```{bash CTX, eval=FALSE, tidy = FALSE}
ctx_args=(
    # adjacency table output from grn
    "${SCENIC_DIR}/pbmc_adj_3k.tsv"
    # gene rankings for motif enrichment within 10 kb of the TSS
    "${SCENIC_DIR}/hg38_10kbp_up_10kbp_down_full_tx_v10_clust.genes_vs_motifs.rankings.feather"
    # gene rankings for motif enrichment 500 bp upstream to 100 bp downstream of the TSS
    "${SCENIC_DIR}/hg38_500bp_up_100bp_down_full_tx_v10_clust.genes_vs_motifs.rankings.feather"
    --annotations_fname "${SCENIC_DIR}/motifs-v9-nr.hgnc-m0.001-o0.0.tbl"   # TF motif annotations
    --expression_mtx_fname "${SCENIC_DIR}/pbmc3k_counts.loom"              # counts matrix loom file
    --mode "dask_multiprocessing"                                           # enable dask multi-core processing
    --output "${SCENIC_DIR}/regulons_pbmc_3k.csv"                           # output file of learned TF regulons
    --num_workers 6                                                         # number of CPU cores
)
singularity exec "${SCENIC_DIR}/aertslab-pyscenic-0.12.1.sif" pyscenic ctx "${ctx_args[@]}"
```

### aucell: calculate TF activity scores

Enrichment of a regulon is measured as the area under the recovery curve (AUC) of the genes that define the regulon.

```{bash AUCELL, eval=FALSE, tidy = FALSE}
aucell_args=(
    "${SCENIC_DIR}/pbmc3k_counts.loom"     # counts matrix loom file
    "${SCENIC_DIR}/regulons_pbmc_3k.csv"   # regulon table output from ctx
    -o "${SCENIC_DIR}/auc_pbmc_3k.csv"     # cell x TF matrix of AUC values
)
singularity exec "${SCENIC_DIR}/aertslab-pyscenic-0.12.1.sif" pyscenic aucell "${aucell_args[@]}"
```

The csv files created in this analysis can be used as the TF activity inputs in a dominoSignal analysis.

## Related articles


- [Get started with dominoSignal](https://fertiglab.github.io/dominoSignal/articles/dominoSignal.html): the basic workflow
- [Interacting with domino objects](https://fertiglab.github.io/dominoSignal/articles/domino_object_vignette.html): object structure and accessor functions
- [Plotting functions and options](https://fertiglab.github.io/dominoSignal/articles/plotting_vignette.html): visualization options
- [Differential signaling workflow](https://fertiglab.github.io/dominoSignal/articles/differential_signaling.html): comparing signaling across subjects
- [Using the CellPhoneDB database](https://fertiglab.github.io/dominoSignal/articles/cellphonedb_vignette.html): preparing a receptor-ligand map

***

To report a bug, ask a question, or suggest a feature, please [open an issue](https://github.com/FertigLab/dominoSignal/issues).

<details><summary>Vignette Build Information</summary>
Date last built and session information:
```{r}
Sys.Date()
sessionInfo()
```
</details>
